Human Factors Influencing Visual Statistical Inference

### 2013-07-24 sociology_paper.tex
typo fixed

Model \eqref{eqn:demographic_response} is fitted to the data with fixed effect factors such as age, country, education and gender. To estimate the overall factor main effect, a reduced model is fitted removing that factor from Model \eqref{eqn:demographic_response}. Analysis of variance (ANOVA)  results shown in Table~\ref{tbl:anova_factor} suggest that all the factor variables are significantly different in describing the the probability of correct response except gender. Gender does not show any significant difference in performance.  Similarly, ANOVA results are obtained by fitting Model \eqref{eqn:demographic_time} to the data of log time taken and are shown in Table~\ref{tbl:anova_factor} as well. All the factor variables are significantly different for log time taken including gender.

@@

To further illustrate how the individual factor levels differ from each other  the detailed results of Models \eqref{eqn:demographic_response} and \eqref{eqn:demographic_time} are shown in Table \ref{tbl:model_result_demographics}. For parameter estimation the first factor levels shown in the Table are  used as the base line. Demographic factor levels have significant effects on log time taken to evaluate a lineup. But not all the levels are significant for the probability of correct responses.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Models \eqref{eqn:demographic_time} and \eqref{eqn:demographic_response} fitted for average log time taken and probability of correct lineup evaluations respectively. For time taken all the demographic factors are significant. For probability of correct response age group 36-40, rest of the world and graduate degree are significantly different. For gender no difference in performance is observed. Lineup variability is estimated to be very large for Model \eqref{eqn:demographic_response}.}
\scalebox{0.86}{
\begin{tabular}{rlrrrrrrrrrr}
  \hline
Demographic&& \multicolumn{4}{c} {Model \eqref{eqn:demographic_time} Log Time} & &\multicolumn{4}{c} {Model \eqref{eqn:demographic_response} Proportion Correct} \\

@@

Notice that lineup specific variance estimate is 5.259 with a standard error of 2.293 which is much higher than the other significant parameter estimates in Model \eqref{eqn:demographic_response}. This indicates that the major and most important factor affecting the probability of correct response is  lineup difficulty. This is also evident from the large variability in the proportion correct in Figure~\ref{fig:demographic_effect}. Because of this huge impact of lineup specific variance, the practical impact of other factor variables on the probability of correct response is in fact very small. We illustrate this with the following example of graduate degree.

@@

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately.  As an alternative to Model \eqref{eqn:trend_response} we examined a reduced model with attempt as a continuous covariate. They did not appear to be significantly different with a $p$-value of 0.856. But we consider the bigger Model \eqref{eqn:trend_response} with attempt as a factor variable with 10 levels to examine the effect of each level. Attempt 1 is considered to be base level while fitting the model.

Table \ref{tbl:model_result_response} presents the parameter estimates and $p$-values of fixed effect estimates of Model \eqref{eqn:trend_response}. The larger $p$-values suggest that none of the levels of attempt ($\alpha_2$ through $\alpha_{10}$) are significant at \%1 significance level. Moreover, some of the estimates are positive and some are negative and they show up seemingly in random order suggesting later attempts not necessarily have improved. This indicates that there may be no learning effect on the probability of correct evaluations.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_response} fitted for probability of correct lineup evaluation. None of the fixed factor effects of attempt ($\alpha_2$ through $\alpha_{10}$) are significantly different from the first attempt $\alpha_1$
at \%1 level in all three experiments 5, 6 and 7. For experiment 7 subject specific variation is very small on the other hand lineup variance is much higher compared to the other two experiments.}
\scalebox{0.81}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

To visualize how the performance in correct response improves over successive attempts, we fitted Model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. Least square regression lines were fitted through the subject specific residuals as shown in Figure~\ref{fig:learning_trend_response}. Two important features were observed; one is subject specific variability and the other is random slope with attempts which indicates subjects specific learning trend. Some subjects show improvement over time and some show the decrease in performance.

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=6.3in]{learning_trend_subject.pdf} 
   \caption{ Least square lines fitted through the subject specific residual proportion correct obtained from Model \eqref{eqn:trend_response} fitted without attempt are plotted against attempt. Subject specific positive and negative slopes are observed. Mean residuals are shown as dots and least square regression lines fitted through the points show no overall learning trend in each of the three experiments.}
   \label{fig:learning_trend_response}
\end{figure}

@@

Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that on an average later attempt took less time for an evaluation. Even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_time} fitted for log time taken to evaluate a lineup. Both fixed effect parameters of Attempt ($\alpha_1$ and $\alpha$) are highly significant for all three experiments 5, 6 and 7.}
\scalebox{0.81}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

The lineup variance of experiment 7 is estimated as $\hat{\sigma}^2_{\ell} = 10.98$ from Model \eqref{eqn:trend_response}. But from Model \eqref{eqn:trend_time} it is estimated to be much smaller in all these experiments. This suggests that the harder lineups not necessarily took more time than easier lineups. The observers spent enough times to evaluate a easy lineup and for difficult lineup they might just give up at some point and provided the feedback.

@@

To visualize how the time taken reduces over the successive attempts, we fitted Model \eqref{eqn:trend_time} excluding the covariate attempt from the model and computed the residuals. Least square regression lines are fitted through the subject specific residuals. Subject specific slopes are much different in each of the three experiments as we see in Figure \ref{fig:learning_trend_time}. Some subjects improved over attempts by taking less time in the later attempts while others got worse. The averages of these residuals for each attempt are plotted as dots. Least square regression lines are fitted to these points excluding the first attempt since for first attempt we fitted an indicator covariate. The downward trends are evident in the plots. All the slopes are highly significant. As expected we observed large positive residuals for each of the experiments for first attempt.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend_time.pdf}    
 \includegraphics[width=6.3in]{learning_trend_time_subject.pdf}    
   \caption{Least square regression lines fitted through the subject specific residuals obtained by  fitting Model \eqref{eqn:trend_time} without covariate attempt. Differences in subject specific slopes are observed. Some of the subjects did worse over successive attempts while others did better. Averages of these residuals are plotted as dots and least square regression lines are fitted to obtain overall trends. For all the three experiments the overall downward slopes are statistically significant which indicates that MTurk workers take less time as they progress through their attempts.}
   \label{fig:learning_trend_time}
\end{figure}

@@

We fitted Model \eqref{manova} to test if the mean performance vectors are similar for different locations. For this we use {\it anova()} function of {\it stats} package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05. This suggest that there may be no difference in location.

\begin{table}[hbtp]
\caption{The summary of the results obtained by fitting MANOVA Model \eqref{manova} using Wilks test.}
\begin{center}
\begin{tabular}{ccccccccc}
  \hline \hline
 Location & && & \multicolumn{3}{c} {Degrees of Freedom}  & F test \\
 \cline{5-7}
 Effect & DF & Wilks & Approx. F & Numerator& Denumerator &Residual & $p$-value\\
  \hline
  Interaction& 3&0.07986 & 0.592 &15&5.92 & 6 & 0.8082 \\ 
  Genotype & 4&0.01689 & 1.722 & 20&14.22 & 8 & 0.1488 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:manova}
\end{table}

We also examined whether the proportion of correct responses differs if the actual plot is on the outer boundary or in the inner locations. The locations 7,8,9,12,13,14 in the lineup are considered inner locations and the rest of the boundary locations are considered to be outer location. As we see in Figure \ref{fig:location_effect}, location 9, 12 are inside for Interaction effect and location 8, 12 are inside for Genotype. It does not show any differences whether the actual plot is inside or outer border of the lineup. We also fitted Model \eqref{manova} with two locations, inner and outer, as covariate and observed no significant differences.

>>>>

Model \eqref{eqn:demographic_response} is fitted to the data with fixed effect factors such as age, country, education and gender. To estimate the overall factor main effect, a reduced model is fitted removing that factor from model \eqref{eqn:demographic_response}. Analysis of variance (ANOVA)  results shown in Table~\ref{tbl:anova_factor} suggest that all the factor variables are significantly different in describing the the probability of correct response except gender. Gender does not show any significant difference in performance.  Similarly, ANOVA results are obtained by fitting model \eqref{eqn:demographic_time} to the data of log time taken and are shown in Table~\ref{tbl:anova_factor} as well. All the factor variables are significantly different for log time taken including gender.

@@

To further illustrate how the individual factor levels differ from each other  the detailed results of models \eqref{eqn:demographic_response} and \eqref{eqn:demographic_time} are shown in Table \ref{tbl:model_result_demographics}. For parameter estimation the first factor levels shown in the Table are  used as the base line. Demographic factor levels have significant effects on log time taken to evaluate a lineup. But not all the levels are significant for the probability of correct responses.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of models \eqref{eqn:demographic_time} and \eqref{eqn:demographic_response} fitted for average log time taken and probability of correct lineup evaluations respectively. For time taken all the demographic factors are significant. For probability of correct response age group 36-40, rest of the world and graduate degree are significantly different. For gender no difference in performance is observed. Lineup variability is estimated to be very large for model \eqref{eqn:demographic_response}.}
\scalebox{0.86}{
\begin{tabular}{rlrrrrrrrrrr}
  \hline
Demographic&& \multicolumn{4}{c} {Log Time (model \ref{eqn:demographic_time})} & &\multicolumn{4}{c} {Proportion Correct (model \ref{eqn:demographic_response})} \\

@@

Notice that lineup specific variance estimate is 5.259 with a standard error of 2.293 which is much higher than the other significant parameter estimates in model \eqref{eqn:demographic_response}. This indicates that the major and most important factor affecting the probability of correct response is  lineup difficulty. This is also evident from the large variability in the proportion correct in Figure~\ref{fig:demographic_effect}. Because of this huge impact of lineup specific variance, the practical impact of other factor variables on the probability of correct response is in fact very small. We illustrate this with the following example of graduate degree.

@@

\subsection{Learning Trend} models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately.  As an alternative to model \eqref{eqn:trend_response} we examined a reduced model with attempt as a continuous covariate. They did not appear to be significantly different with a $p$-value of 0.856. But we consider the bigger model \eqref{eqn:trend_response} with attempt as a factor variable with 10 levels to examine the effect of each level. Attempt 1 is considered to be base level while fitting the model.

Table \ref{tbl:model_result_response} presents the parameter estimates and $p$-values of fixed effect estimates of model \eqref{eqn:trend_response}. The larger $p$-values suggest that none of the levels of attempt ($\alpha_2$ through $\alpha_{10}$) are significant at \%1 significance level. Moreover, some of the estimates are positive and some are negative and they show up seemingly in random order suggesting later attempts not necessarily have improved. This indicates that there may be no learning effect on the probability of correct evaluations.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of model \eqref{eqn:trend_response} fitted for probability of correct lineup evaluation. None of the fixed factor effects of attempt ($\alpha_2$ through $\alpha_{10}$) are significantly different from the first attempt $\alpha_1$
at \%1 level in all three experiments 5, 6 and 7. For experiment 7 subject specific variation is very small on the other hand lineup variance is much higher compared to the other two experiments.}
\scalebox{0.81}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

To visualize how the performance in correct response improves over successive attempts, we fitted model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. Least square regression lines were fitted through the subject specific residuals as shown in Figure~\ref{fig:learning_trend_response}. Two important features were observed; one is subject specific variability and the other is random slope with attempts which indicates subjects specific learning trend. Some subjects show improvement over time and some show the decrease in performance.

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=6.3in]{learning_trend_subject.pdf} 
   \caption{ Least square lines fitted through the subject specific residual proportion correct obtained from model \eqref{eqn:trend_response} fitted without attempt are plotted against attempt. Subject specific positive and negative slopes are observed. Mean residuals are shown as dots and least square regression lines fitted through the points show no overall learning trend in each of the three experiments.}
   \label{fig:learning_trend_response}
\end{figure}

@@

Table \ref{tbl:model_result_time} presents the results of model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that on an average later attempt took less time for an evaluation. Even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of model \eqref{eqn:trend_time} fitted for log time taken to evaluate a lineup. Both fixed effect parameters of Attempt ($\alpha_1$ and $\alpha$) are highly significant for all three experiments 5, 6 and 7.}
\scalebox{0.81}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

The lineup variance of experiment 7 is estimated as $\hat{\sigma}^2_{\ell} = 10.98$ from model \eqref{eqn:trend_response}. But from model \eqref{eqn:trend_time} it is estimated to be much smaller in all these experiments. This suggests that the harder lineups not necessarily took more time than easier lineups. The observers spent enough times to evaluate a easy lineup and for difficult lineup they might just give up at some point and provided the feedback.

@@

To visualize how the time taken reduces over the successive attempts, we fitted model \eqref{eqn:trend_time} excluding the covariate attempt from the model and computed the residuals. Least square regression lines are fitted through the subject specific residuals. Subject specific slopes are much different in each of the three experiments as we see in Figure \ref{fig:learning_trend_time}. Some subjects improved over attempts by taking less time in the later attempts while others got worse. The averages of these residuals for each attempt are plotted as dots. Least square regression lines are fitted to these points excluding the first attempt since for first attempt we fitted an indicator covariate. The downward trends are evident in the plots. All the slopes are highly significant. As expected we observed large positive residuals for each of the experiments for first attempt.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend_time.pdf}    
 \includegraphics[width=6.3in]{learning_trend_time_subject.pdf}    
   \caption{Least square regression lines fitted through the subject specific residuals obtained by  fitting model \eqref{eqn:trend_time} without covariate attempt. Differences in subject specific slopes are observed. Some of the subjects did worse over successive attempts while others did better. Averages of these residuals are plotted as dots and least square regression lines are fitted to obtain overall trends. For all the three experiments the overall downward slopes are statistically significant which indicates that MTurk workers take less time as they progress through their attempts.}
   \label{fig:learning_trend_time}
\end{figure}

@@

We fitted model \eqref{manova} to test if the mean performance vectors are similar for different locations. For this we use {\it anova()} function of {\it stats} package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05. This suggest that there may be no difference in location.

\begin{table}[hbtp]
\caption{The summary of the results obtained by fitting MANOVA model \eqref{manova} using Wilks test.}
\begin{center}
\begin{tabular}{ccccccccc}
  \hline \hline
 Location & && & \multicolumn{3}{c} {Degrees of Freedom}  & F test \\
 \cline{5-7}
 Effect & DF & Wilks & Approx. F & Numerator& Denumerator &Residual & $p$-value\\
  \hline
  Interaction& 3&0.07986 & 0.592 &15&5.92 & 6 & 0.8082 \\ 
  Genotype & 4&0.01689 & 1.722 & 20&14.22 & 8 & 0.1488 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:manova}
\end{table}

We also examined whether the proportion of correct responses differs if the actual plot is on the outer boundary or in the inner locations. The locations 7,8,9,12,13,14 in the lineup are considered inner locations and the rest of the boundary locations are considered to be outer location. As we see in Figure \ref{fig:location_effect}, location 9, 12 are inside for Interaction effect and location 8, 12 are inside for Genotype. It does not show any differences whether the actual plot is inside or outer border of the lineup. We also fitted model \eqref{manova} with two locations, inner and outer, as covariate and observed no significant differences.
